Subglacial and subaerial fluvial sediment transport capacity respond differently to water discharge variations

### 2024-08-05 Delaneyetal2024.tex
mauro's edits

Despite the different physical processes between the subaerial and subglacial systems, observed water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,hodson1998,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of these studies discuss the disparity between water discharge and sediment export given the role of sediment access or production, along with the sensitive and non-linear relationship between the two.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited regime, where sediment transport responds to the channel's hydraulic conditions, instead of sediment availability.
The response of sediment transport capacity to water discharge variations may affect the interpretation of sediment transport records in these glacierized catchments \citep[e.g.][]{ganti2016,mancini2023}.
This relationship must be clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019b} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

This manuscript has two objectives: 1) to evaluate the variability of  sediment transport capacity in subglacial channels with evolving channel size and different hydrographs compared to subaerial ones, and 2) to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems.
We use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
We run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
Lastly,  we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, steady-state R-channel, and pipe flow conditions.
These formulations extend the relationships presented in \citet{alley1997} and illustrate the sediment transport capacity behavior of different channel types.
Findings indicate differences in the relationships amongst water discharge, channel geometry, and water velocity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

\ian{redo some of this with Mauro's comments}

@@

The first experiment aims to characterize the variability in sediment discharge capacity in subglacial relative to subaerial channels across a range of channel slopes and shapes, along with friction factors.
Additionally, we examine differing water discharge variability by smoothing the water discharge over different periods from $15$\,\unit{min} up to $15$\,\unit{days}.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To evaluate variability, we subtract the model outputs from their values smoothed using daily averages,  detrending the seasonal signal.
We then calculate the variability in the model output by taking the standard deviation of this detrended time series.

The effects of channel shape, gradient, and roughness are established by running the model with random parameter values of channel slope and geometry factors ($\beta$,$\nabla z_c$, and $\alpha$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.5$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$ \citep[e.g.][]{werder2010b,magnusson2012,chandler2013}.
To be accepted, subglacial model runs must experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.
The initial condition of cross-sectional area $S_g$ is evaluated by applying a model spinup with repeatedly applying the maximum observed water discharge of the first $4$ days of the study until there is no change in the channel area.

@@

In the second experiment, the models are applied to a reference test case for each glacier.
These experiments assumes  a subglacial channel with $\beta=\frac{\pi}{6}$ and a subaerial channel with $\alpha = \frac{1}{3}$ and slope of $0.02$ (Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities \citep[$\sim\,1.6\,$\unit{m}\,\unit{s}$^{-1}$][]{werder2010b,chandler2013} occur for both \alpine{} and \icesheet{}.
% 
To test the covariance between sediment transport capacity and water discharge, model outputs are compared to water discharge from the glaciers using Spearman rank correlation.
This metric accounts for the ordering of values, but not their magnitude.
Rank correlation reduces the impact of the non-linear relationship between sediment transport capacity and hydrology.

@@

As in Section~\ref{sect:fluv}, the subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4h$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$.
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady-state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the Shreve \citeyear{shreve1972} potential and that $S$ adjusts in steady state with $\Psi$ and $Q$.
Note that the R-channel model used above (Section~\ref{sect:sub_mode}), Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time-evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow-like conditions occur when an R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both steady-state R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

@@

R-channels rarely operate in a steady state with variations in water discharge, especially during severe rain or melt is too short for the channel to reach a steady state (Figures~\ref{fig:model_outs} and \ref{fig:Qw_vari}).
In these cases with high water discharge variability, channels can behave more like a pipe of fixed cross-section \citep[\alpine{} case in Figure \ref{fig:Qw_vari}; e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two channel types when that exponent is at most $\frac{5}{3}$.
Thus fluctuations of discharge on short timescales (on the order of a day; Figure~\ref{fig:multi_run}) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep[][]{alley1997}.
Note that pipe flow would cause sediment discharge capacity to covary with water discharge (Figure~\ref{fig:chan_types} a).
Covariance also occurs for steady state R-channels, but not for R-channels that evolve in time . (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types} b).

@@

Several aspects of the models and experiment design make the comparison of the subaerial and subglaical systems difficult. 
The lumped nature of the models means that they operate independently of the upstream drainage network.
Additionally, we omit analysis of suspended sediment discharge records due to their dependence on sediment supply in capturing variations in sediment discharge, in addition to sediment transport capacity \citep[e.g.][]{delaney2019}.
In reality, processes such as sediment access are highly important in controlling sediment export in glacierized catchments \citep[e.g.][]{herman2015,vergara2022}.
Furthermore, meltwater can be distributed to flow through several adjacent conduits impacting the sediment transport capacity in each \citep[e.g.][]{werder2013,hewitt2019,delaney2023}.
The lumped models used here isolate the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems. 
However, they neglect more complex, yet important spatially distributed processes.

@@

The manuscript's first objective aims to evaluate the relative variability in sediment transport capacity between subglacial and subaerial channels.
Results demonstrate that sediment transport capacity variability is higher in pressurized subglacial channels independent of hydrological conditions and channel characteristics.
Increased variability occurs even in environments where the water discharge has relatively small diurnal variations in the \icesheet case.
Yet, elevated variability is especially strong with greater water discharge variations in the \alpine case.
Further evaluation is needed to establish the role of high variability in sediment transport capacity in evaluating hydro-climatic signals from sediment records.
Greater variability in sediment transport capacity may also lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.

The manuscript's second objective is to establish if water discharge covaries with subglacial sediment transport capacity.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during an discharge event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most systems.
This incoherent relationship presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Covariance between water discharge and sediment transport capacity could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case.

>>>>

Despite the different physical processes between the subaerial and subglacial systems, observed water discharge and sediment export are often compared in contemporary glacierized catchments \citep[e.g.][]{willis1996,hodson1998,pearce2003,richards2003,swift2005,chu2009,tedstone2012,chu2012,overeem2017,delaney2018,swift2021,lu2022,andresen2024}.
Many of these studies discuss the disparity between water discharge and sediment export, especially given the role of sediment access or production.
Yet, a variable relationship between water discharge and sediment export could be expected from glaciers with evolving conduit size in a transport-limited regime, where sediment transport responds to the channel's hydraulic conditions, instead of sediment availability.
The response of sediment transport capacity to water discharge variations may affect the interpretation of sediment transport records in these glacierized catchments \citep[e.g.][]{ganti2016,mancini2023}.
This relationship must be clarified as sediment dynamics change in glacierized regions along with changing hydrology \citep[e.g.][]{brunner2019}, increasing variability of water discharge \citep{lane2019b} and occurrence of extreme glacier melt  \citep[e.g.][]{overeem2015,cremona2023} that may drive high magnitude sediment transport events.

This manuscript has two objectives: 1)  to establish whether sub-seasonal water discharge can co-vary with sediment transport capacity in subglacial systems, and 2) to evaluate the variability of  sediment transport capacity in subglacial channels with evolving channel size and different hydrographs compared to subaerial ones.
We use numerical models to examine sediment transport capacity in both subglacial channels that evolve in size and subaerial channels that remain fixed.
They are applied to proglacial hydrological records from an Alpine glacier in Switzerland (Fieschergletscher) and a land-terminating glacier in Greenland (Leverett Glacier).
Model outputs demonstrate the specific processes that can drive variability in sediment discharge capacity in subglacial systems.
We also run a model ensemble to evaluate the sediment transport capacity variability in subglacial channels compared to subaerial channels.
Lastly, we present algebraic formulations of the sediment transport capacity response to water discharge in subaerial, steady-state R-channel, and pipe flow conditions.
These formulations extend the relationships presented in \citet{alley1997} and illustrate the sediment transport capacity behavior of different channel assumptions.
Findings indicate differences in the relationships amongst water discharge, channel geometry, water velocity and sediment transport capacity in subglacial and subaerial channels.
The manuscript then discusses the implications of variability in sediment transport capacity from glaciers and the interpretation of sediment transport records.

@@

@@

In the first experiment, the models are applied to a reference test case for each glacier.
These experiments assumes  a subglacial channel with $\beta=\frac{\pi}{6}$ and a subaerial channel with $\alpha = \frac{1}{3}$ and slope of $0.02$ (Table~\ref{table:vpm}).
Both friction factors $f_r$ and $f_f$ are tuned so that reasonable water velocities \citep[$\sim\,1- 1.5\,$\unit{m}\,\unit{s}$^{-1}$][]{werder2010b,chandler2013} occur for both \alpine{} and \icesheet{}.
To test the covariance between sediment transport capacity and water discharge, model outputs are compared to water discharge from the glaciers using Spearman rank correlation.
This metric accounts for the ordering of values, but not their magnitude.
Rank correlation reduces the impact of the non-linear relationship between sediment transport capacity and hydrology.

The second experiment aims to characterize the variability in sediment discharge capacity in subglacial relative to subaerial channels across a range of channel slopes and shapes, and friction factors.
Additionally, we examine the effects differing water discharge smoothing periods from $15$\,\unit{min} up to $15$\,\unit{days}.
The results below present water velocity ($v_g$, $v_f$), shear stress ($\tau_g$, $\tau_f$), and width-integrated shear stress ($w_g\tau_g$, $w_f\tau_f$) from the subglacial and subaerial models.
For brevity, these values together are referred to as ``model outputs'' in the text.
To evaluate variability, we subtract the model outputs from their daily averages. 
This creates a time series that is detrended from he seasonal variations in the model outputs and has an approximately normal distribution.
We then present the variability in the model output by taking the standard deviation of this detrended time series.

The effects of channel shape, gradient, and roughness are established by running the model with random parameter values of channel slope and geometry factors ($\beta$,$\nabla z_c$, and $\alpha$) and friction factors ($f_r$ and $f_f$; see Table~\ref{table:vpm} for value range).
Runs with parameter combinations are accepted if their mean subglacial water velocity over the season lies between $0.5$\,\unit{m}\,\unit{s}$^{-1}$ and $2$\,\unit{m}\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}\,\unit{s}$^{-1}$ \citep[e.g.][]{werder2010b,magnusson2012,chandler2013}.
To be accepted, subglacial model runs must experience a flotation fraction ($\frac{\Delta h}{2\,h_o}$) of above $1.2$ for less than $2.5$\,\% of the run.
A model spinup dictates the initial condition of cross-sectional area $S_g$.
The spinup consists of applying the maximum observed water discharge of the first $4$ days of the study period to the model until there is no change in the channel area $S_g$.

@@

@@

As in Section~\ref{sect:fluv}, the subaerial channel is assumed to have a width $w$ much greater than its depth $H$, such that $D_h\approx 4H$, and to have a constant head gradient ($\Psi$) given by the topography.
Further, it is assumed that its width can be approximated by a relation $w \propto Q^\alpha$ (Equation~\ref{eq:wcf}) with $\alpha\in [0,1]$.
End members $\alpha=0$ or $\alpha=1$ correspond a  subaerial channels of constant width (a slot canyon) or depth (no natural equivalent), respectively.
% 
For a steady-state R-channel, it is assumed that  $\Psi$ is constant (approximated by the gradient of the \citet{shreve1972} potential) and that $S$ adjusts in steady state with $\Psi$ and $Q$.
Note that the R-channel model used above (Section~\ref{sect:sub_mode}), Equation~\ref{eq:dS_dt} calculates $\Psi$ from the time-evolving $S_g$ via the Darcy-Weisbach Equation~\ref{eq:dh} and thus no Shreve approximation is then needed.
% Conversely, pipe flow has $S$ fixed and $\Psi$ adjusts to satisfy the given $Q$.
% 
Pipe flow-like conditions occur when an R-channel is subjected to rapid discharge variations such that the channel cannot adjust its size.
In this case, it is assumed that the cross-sectional area $S$ is fixed, and $\Psi$ adjusts to the specified $Q$.
For both steady-state R-channel and pipe flow, it is assumed that $D_h \propto S^{1/2}$.

@@

R-channels rarely operate in a steady state with variations in water discharge, especially during severe rain or melt is too short for the channel to reach a steady state (Figures~\ref{fig:model_outs} and \ref{fig:Qw_vari}).
In these cases with high water discharge variability, channels can behave more like a pipe of fixed cross-section \citep[\alpine{} case in Figure \ref{fig:Qw_vari}; e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two channel types when that exponent is at most $\frac{5}{3}$.
Thus fluctuations of discharge on short timescales (on the order of a day; Figure~\ref{fig:multi_run}) have the potential to cause conditions with very high sediment transport capacities.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep[][]{alley1997}.
Note that pipe flow would cause sediment discharge capacity to covary with water discharge (Figure~\ref{fig:chan_types} a).
Covariance also occurs for steady state R-channels, but not for R-channels where channel area evolves in time (Figures~\ref{fig:Qw_vari} and \ref{fig:chan_types} b).

@@

Several aspects of the models and experiment design make the comparison of the subaerial and subglaical systems difficult. 
The lumped nature of the models means that they operate independently of the upstream drainage network.
Additionally, we omit analysis of suspended sediment discharge records due to their dependence on sediment supply in capturing variations in sediment discharge, in addition to sediment transport capacity \citep[e.g.][]{delaney2019}.
In reality, processes such as sediment access are usually important in controlling sediment export in glacierized catchments \citep[e.g.][]{herman2015,vergara2022}.
Furthermore, meltwater can be distributed to flow through several adjacent conduits impacting the sediment transport capacity in each \citep[e.g.][]{werder2013,hewitt2019,delaney2023}.
The lumped models used here isolate the relationship between water discharge and sediment transport capacity in subglacial and subaerial systems. 
However, they neglect more complex, yet important, spatially distributed processes.

@@

The manuscript's first objective is to establish if water discharge covaries with subglacial sediment transport capacity.
In subglacial channels, the timing of peak sediment discharge capacity occurs before peak water discharge during an discharge event, due to evolving channel size.
In subaerial channels, the timing of peak sediment discharge capacity and water discharge coincide.
Results here suggest that, even in a transport-limited subglacial system, an incoherent relationship between water and subglacial sediment discharge could be expected in most systems.
This incoherent relationship presents a challenge in linking hydro-climatic conditions or events to sediment export from glaciers.
Water discharge and sediment transport capacity covariance could be possible when water discharge varies at a slower rate than subglacial channel size.
In natural environments, this appears to rarely be the case.

The manuscript's second objective aims to evaluate the relative variability in sediment transport capacity between subglacial and subaerial channels.
Results demonstrate that sediment transport capacity variability is higher in pressurized subglacial channels that are out of equilibrium with water discharge.
Increased variability occurs even in environments where the water discharge has relatively small diurnal variations in the \icesheet case.
Yet, elevated variability is especially strong with greater water discharge variations in the \alpine case.
Further evaluation is needed to establish the role of high variability in sediment transport capacity in evaluating hydro-climatic signals from sediment records.
Greater variability in sediment transport capacity may also lead to sediment exhaustion by subglacial water repeatedly crossing the mobilization threshold across a range of water discharges.
